Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-10 paper.Rmd
copy edits
I'm sure I missed some things in this reading. I'll pick it up again tomorrow

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkable accurately.'' Indeed they argue that this so-called theoretical model outperforms four other ``more mechanical'' alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

The corresponding results for each question are shown in B&S Tables 1 and 2, which we will examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. 
The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

@@

Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case. There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to ...

@@

Far from being a trivial choice, the question of whether to use ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results. B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.

@@

> The final approach is a random forest analog to split-population modeling.
first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).

The intention of B&S appears to have been to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models. The models are also identical otherwise, i.e. they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their research strategy.

@@

By examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

This implies that the ROC curves were only smoothed in the referenced Figure 1, but actually all AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result appear step-like. Figure 1 shows our replication of both the smoothed ROC curves B&S report, and the actual ROC curves on the right. The standard method is to compute the area under the curve AUC statistic on the original, non-smooth ROC curves that are shown on the right.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results: the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.

@@

1. B&S: ``Of the approaches we test, the split-population analog is most promising \ldots ''

@@

2. B&S: ``Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots ''

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a maring in the 6-month data version.

3. B&S: ``The weighted model performs very poorly regardless \ldots ''

@@

4. B&S: ``Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent. [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take. \ldots ''

@@

tab4o <- tab4o %>%
  rename(header = type) %>%
  mutate(source = "Original, scored with civil war incidence")
tab4f <- tab4f %>%
  rename(Predicted0 = `0`, Predicted1 = `1`) %>%
  mutate(source = "Fixed, scored with civil war onset")

@@

The B&S advocate the use theory to guide prediction. But theory as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

>>>>

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkable accurately." Indeed they argue that this so-called theoretical model outperforms four other "more mechanical" alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust, and a more precise test is undertaken. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

The corresponding results for each question are shown in B&S Tables 1 and 2, which we examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values, or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

@@

Although smoothing might appear to be a simple option to aid in the visual interpretation of ROC curves, and is presented as such by B&S, it actually involves estimating an underlying population ROC curve, using a parametric binormal model in this case. There is a literature in statistics and the medical field on such ROC estimators and the conditions under which they might be preferable to empirical ROC curves. For example, when the goal is to extrapolate from a statistical sample to a larger population of potential cases...

(RM: IDK. But I think it can be short and simple. Highlight that smoothing in unconventional in conflict/political forecasting)

@@

Far from being a trivial choice, the question of whether to use smoothed ROC estimates rather than the empirical ROC curves is a major decision that dramatically impacts the results. B&S do not offer a justification for departing from the common practice of using the empirical ROC curves.

@@

> The final approach is a random forest analog to split-population modeling. We first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.^[Disentangling this coding error is not straightforward as it occurs over several R scripts and requires (or at least is easier to verify by) running partway through the actual replication until the objects holding the training data for the models are instantiated and can be examined. We have documented details at https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5.]

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

However, by examining the code used to generated Table 4, we were able to confirm that the onset forecasts are assessed using incidence, not onset. In the file `6mo_make_confusion_matrix.do` on line 52, missing values in "incidence_civil_ns" are recoded to 1, thus reverting the onset coding of this variable to incidence.

@@

This implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure 1 shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

In sum, it is not only the case that using smoothed ROC curves alters the results, but also that the use of smoothed ROC curves to calculate AUC-ROC values benefits \textbf{only} the escalation model. It does so consistently and by a considerable margin. All eight alternative models reported in B&S Tables 1 and 2, on average, do not gain when using smoothed ROC curves to calculate AUC.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.

@@

1. B&S: "Of the approaches we test, the split-population analog is most promising \ldots "

@@

2. B&S: "Adding PITF predictors improves the performance of the escalation model over six-month windows but diminishes it over one-month windows \ldots "

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a margin in the 6-month data version.

3. B&S: "The weighted model performs very poorly regardless \ldots "

@@

4. B&S: "Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent. [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take. \ldots "

@@

tab4o <- tab4o %>%
  rename(header = type) %>%
  mutate(source = "Original, scored with civil war incidence")
tab4f <- tab4f %>%
  rename(`Predicted 0` = `0`, `Predicted 1` = `1`) %>%
  mutate(source = "Fixed, scored with civil war onset")

@@

B&S advocate the use theory to guide prediction. But theory as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.
